Asymmetric relationships and their effects on coexistence

### 2023-07-28 paper/Paper.Rnw
added an Intro para on intuition behind ATAs

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argue that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} show that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples have also been previously explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

@@

%Intro to storage effects and why they might be influenced by ATAs
It is possible to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular, storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs 
may be relevant to them. Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on positive covariation between the fluctuating quality of the environment
for a species (E) and the strength of competition (C) experienced by that species (called EC covariance);
and on the assumption that environmental responses for distinct species are also distinct,
so that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive EC covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent EC covariance contributes to 
differences between currently rare and currently common species growth rates \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, 
and ATAs, when they occur between environment and competition variables, can be an aspect of EC covariation.

@@

>>>>

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study such relationships, ecologists 
very commonly use metrics such as correlation and covariance. But these standard measures of the strength 
and direction of association between variables often do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c all have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has previously been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations and covariance are also often used (e.g., various forms of 
regression), but correlations are common, and other common methods also reveal only a small portion 
of the information content of the relationship between variables \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently several 
insights have been gained by studying how features of associations beyond standard measures 
impact various ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting the potential for rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs in nature using 
plankton population density time series; many other examples have also been previously explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition and coexistence have not been investigated. Doing so is the goal of this study.

@@

%Intro to storage effects and why they might be influenced by ATAs
It is possible to see why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs 
may be relevant to them. Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species (E) and the strength of competition (C) experienced by that species (called EC covariance);
and on the assumption that environmental responses for distinct species are also distinct,
so that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive EC covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent EC covariance contributes to 
differences between currently rare and currently common species growth rates \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, 
and ATAs, when they occur between environment and competition variables, can be an aspect of EC covariation.

@@

%Some intuition
We next give the beginnings of some intuition about how ATA effects work, with a fuller understanding
necessarily coming only later in the paper.
It is known that storage effects can occur if some buffering mechanism reduces losses in bad years so that
gains in good years can more than make up for them. 
%
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, as well as 2) when $E$ is very unfavorable,
or moderately unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is very favorable, or moderately favorable and $C$ is not too strong,
as well as 2) when $E$ is moderately unfavorable but $C$ is very weak. 
%
It turns out that the potential for storage to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal.
Depending on the nature of population growth, storage potential may differ
across the spectrum from 1 to 2.
%
ATA effects are essentially the extent to which asymmetries of association between $E$ and 
$C$ are aligned with the differences in the potential for storage across the dichotomy 1 versus 2.
If $E$ and $C$ are strongly associated in the parts of their distribution where
storage potential is high, ATA effects will be relatively low, and vice versa.
We spend most of the paper formalizing ATA effects in MCT and demonstrating to what 
extent ATA effects can be strong versus weak, but we return in the Discussion to the task
of building an intuition for how ATAs work.
